%************************************************
\chapter{Introduction}\label{ch:introduction}
%************************************************
\section{Subject Matter}\label{sec:subject matter}
%%%
Data, information and knowledge are the basis of all interaction, for example, the diagnosis of a patient or treatment documentation by a physician, in any area of healthcare.
Their reliable, efficient and intuitive access, management, storage, processing and provision is therefore crucial to provide a high quality of practice in all healthcare institutions and education.
%

%
Medical informatics plays a decisive role in this by developing and implementing efficient information systems and technologies within these institutions.
Thereby using theories and methods, procedures and techniques from computer science and other sciences to contribute to the best possible healthcare.
%
In educational contexts such as universities, the basic practical and theoretical approach of students and teachers to knowledge of domains such as medical informatics is mainly developed through the literature, such as \citet{bb2}.
%

%
Various approaches to reliable, efficient and intuitive knowledge retrieval and processing in the field of medical informatics without having to manually read a textbook, such as \citet{bb2}, have been researched and developed.
%
One such initiative is the \ac{snik}-project and the ontology \ac{snik} contained therein.
Both were developed at the Institute for Medical Informatics, Statistics and Epidemiology (IMISE) of the University of Leipzig.
\ac{snik} contains and provides concentrated knowledge from various sources.
Thus, it allows efficient access to technical terms, as well as to roles and their functions and interconnected dependencies in healthcare institutions.
%

\textit{Efficient and reliable access} to knowledge is of significant relevance in educational and research scenarios.

For example, when preparing for exams or academic discussions.
The ability to quickly and intuitively process and retrieve relevant knowledge through available applications is therefore essential.
Furthermore, this ability should be complemented by a reproducible, uniform and comparable evaluation of the results of such applications.
%

%
\acp{llm}, such as GPT-$5$ (ChatGPT), are recent advancements in the field of artificial intelligence.
They are able to analyse and process large amounts of information and data.
They can then provide, for example, insight into the information and knowledge contained therein based on user queries.
%

The resulting possibilities are transformative for how we learn, research, and communicate knowledge.
%
\acp{llm} can help overcome challenges such as identifying, understanding and explaining complex contextual relationships within scientific literature.
At the same time, they can provide everyone an intuitive access to knowledge without time restrictions.
%
To ensure that these possibilities for knowledge retrieval and processing using \acp{llm} provide and meet individual requirements, such reliability and efficiency, a structured, repeatable, and reproducible evaluation is necessary. 
Only through such an assessment can different approaches be objectively compared.
%
\section{Problem Statement}\label{sec:problem_statement}

The practical application and further development of existing systems, for example in medical institutions, requires a uniform understanding of underlying concepts, structures and processes.
%

%
Students and practitioners of medical information systems and their management, as part of the present and future of these systems, need a fundamental knowledge base for this uniform understanding.
In particular, they need fast, intuitive and reliable access to the knowledge contained therein.
%
The corresponding literature, such as \citet{bb2} in the field of medical informatics, representing the knowledge base, is often extensive.
Relevant information and knowledge are fragmented and dispersed in multiple sections and chapters.
Thereby, hindering a fast and intuitive retrieval and synthesis of complex information and knowledge.
%

%
In order to fully grasp individual concepts and their interrelationships, students and practitioners often have to manually read large sections of a book.
Identifying relevant information and knowledge for specific and individual problems is tedious, time-consuming and prone to incompleteness.
%
For example, if a student needs a clear definition of a role in hospital management and the additional functions and responsibilities associated with this role, relevant information is often mentioned and explained in various sections of \citet{bb2}.
%
It is therefore difficult for students and practitioners to acquire knowledge from \citet{bb2} quickly, intuitively and completely at any time.
%

%
Considerable advances, as mentioned in \cref{sec:motivation}, have been achieved in the development of methodologies for knowledge retrieval and processing in medical informatics. 
%
At the same time, the retrieved information and knowledge is either factually incomplete or not aligned with the original source ~\citep[cf.]{Paul_Keller}.
Furthermore, it does not provide a coherent, well-structured, and above all, automatically generated explanation of the requested concept, fact or context ~\citep[cf.]{snikquiz}, \citep[cf.]{hannesbell}.
%

%
\acp{llm} are recent and significant advancements in artificial intelligence and information retrieval.
They allow whole books, such as \citet{bb2}, as a large context input, within a user query.
They further provide comprehensive analytical and reasoning abilities based on that input and their internal knowledge.
%
When using \acp{llm} to retrieve knowledge from \citet{bb2}, an automatic, objective and repeatable evaluation and comparison of the generated results is required.
However, this is not available for students and practitioners of medical informatics.
%
Key problem:
\begin{itemize}
\item \textbf{Problem 1:}\ac{llm}-based knowledge retrieval for \citet{bb2} lacks a standardized and reproducible evaluation, hindering objective assessment and comparison of its capabilities.
\end{itemize}
%
  %\item \textbf{Problem 2:} Fragmentation of knowledge still makes it difficult for students to intuitively and effectively extract coherent and relevant information from \citet{bb2}.
%\end{itemize}
%
Addressing these challenges is essential for enhancing the accessibility and utility of domain-specific knowledge, ultimately supporting both learning and research activities.
%
\section{Motivation}\label{sec:motivation}
 
Previous projects in the context of medical informatics, such as the \ac{snik} ontology, solved crucial aspects of knowledge extraction that are relevant to these demands.
Through the development of \ac{snik}, knowledge from sources on health information systems and their management (HIM) has been accumulated, processed and transformed into a machine-accessible, networked resource.
%
\citet{hannesbell, hannesbell_skill, qanswer} exemplify through practical applications the considerable possibilities of \ac{snik} for more intuitive knowledge retrieval.
They are, however, limited when processing complex queries, cross-linked information scattered across sections and the automatic generation of answers that meet individual requirements.

%
To further address these limitations and the fragmentation of domain-specific knowledge in \cite{bb2}, advanced \acp{llm} promise a more intuitive, on-demand access to complex information and knowledge.
Individually fine-tuning an open-source \ac{lm}, such as Llama-$2$, has demonstrated improved performance over untrained versions.
However, still showing significant performance differences compared to more powerful \acp{llm}, such as \textit{OpenAI's GPT-$4$}.

In conclusion, \citep{Paul_Keller} suggests that more capable \acp{llm} models should be used to improve and overcome these deficiencies in knowledge extraction and processing.
Furthermore, the work of \citet{Paul_Keller} provides comparable results to make current and subsequent approaches to the application of \ac{llm} for knowledge extraction from \citet{bb2} objectively comparable for users.
%

%
Continuous adaptation and advancements of \ac{llm} in knowledge retrieval contribute to overcome the problem of intuitively and quickly extracting fragmented knowledge.
These \acp{llm} can provide an automatically generated explanation of the requested concept, fact or context with a high level of factual reliability.
%

%
The key challenge is therefore to automatically and systematically evaluate \ac{llm}-based knowledge retrieval approaches for \citet{bb2} as a knowledge base.
This will enable an objective evaluation and comparison of different retrieval methods.
This evaluation then serves as the foundation for understanding how these approaches can be leveraged to provide accurate, intuitive, and reliable access to domain-specific knowledge.
%
\section{Goals}\label{sec:goals}
The following goals of this work are assigned to the problems shown in \cref{sec:problem_statement}.
\begin{itemize}
  \item goal G1\label{def:goal 1}:\\
    A reproducible open-source application that integrates selected \ac{lm} for automatic, efficient and intuitive \ac{qa} from \citet{bb2}.
    The application enables an automatic evaluation of the generated results.
    The application has to be accessible and executable with limited resources and available at any time.
   \item goal G2\label{def:goal 2}:\\
    Automatic answering of questions about information systems in healthcare and their management using \citet{bb2} as a knowledge base.
  \item goal G3\label{def:goal 3}:\\
   Solving a sample exam of the module \enquote{Architecture of Information Systems in Healthcare}\footnote{\raggedright{}a module of the Master's program in Medical Informatics at the University of Leipzig, which is based on \citet{bb2}} with the help of \acp{llm}.
   \item goal G4\label{def:goal 4}:\\
   A comparable evaluation of the results provided by the \acp{llm} used in the application based on the metrics used in \citet{Paul_Keller}.
\end{itemize}
\section{Task}
\begin{itemize}
  \item Tasks for Goal G1
        \begin{itemize}
          \item Task A1.1:\label{task:1_1_llm_comparison}\\
            Current \ac{lm} must be selected and compared based on an analysis of their availability and accessibility.
            Within the scope of this work, no claim of completeness can be made regarding these models.
          \item Task A1.2:\label{task:1_2_reproducible}\\
            Setup an open-source application that is reproducible and executable by resource-constrained users.
          \item Task A1.3:\label{task:1_3_answer_generation_llm}\\
            Enable automatic generation of answers via selected \ac{llm} to the predefined questions from \citet{Paul_Keller}.  
          \item Task A1.4:\label{task:1_4_automatic_eval}\\
            Enable automatic evaluation of the generated answers due to the predefined criteria from \citet{Paul_Keller}.
          \item Task A1.5:\label{task:1_5_intuitive_comparison}\\
            Enable an intuitive comparison of the selected \ac{llm}.
        \end{itemize}
  \item Task for goal G2
        \begin{itemize}
            \item Task A2.1:\label{task:2_1_auto_generate}\\
            The selected \ac{lm} integrated in the application are then used to automatically answer the questions provided by \citet{Paul_Keller}.
            This means that the understanding of the question, as well as the identification and retrieval of important knowledge from \citet{bb2} is carried out solely by the \acp{lm}.
            \item Task A2.2:\label{task:2_2_processing}\\
            Processing of the generated answers for further evaluation.
        \end{itemize}
 \item Task for target G3
        \begin{itemize}
        \item Task A3.1:\label{task:3_1_exam}\\
            Automatically solving the exam using the application.
        \item Task A3.2:\label{task:3_2_processing_results}\\
            Processing of the generated results for further evaluation.
        \end{itemize}
 \item Tasks for target G4
    \begin{itemize}
         \item Task A4.1:\label{task:4_1_automatic_eval}
            Automatically evaluate the generated answers in response to the questions according to the same criteria as in \citet{Paul_Keller}.   
          \item Task A4.2:\label{task:4_2_automatic_eval_test}
          Evaluation of the test answers generated by the \ac{llm}  
          \item Task A4.3:
            Provide a comparison of the models and methods.
            This allows a statement to be made about the possible use of these language models in an academic and scientific environment.
        \end{itemize}
\end{itemize}
%
\section{Structure of the thesis}

Chapter 1 describes the basic context of this thesis, identifies existing problems and requirements, proposes objectives for resolving these problems, and outlines tasks for implementing these objectives.
Chapter 2 lays the groundwork for understanding the technology used in this thesis, whilst Chapter 3 summarises the current state of research.
Chapter 4 outlines solution strategies for the problem formulated in ~\cref{sec:problem_statement}, followed by a description of the implementation of these solutions in Chapter 5.
The results of this thesis are presented in Chapter 6 and discussed in summary in Chapter 7.
Chapter 8 concludes with an outlook for this thesis.
