%*****************************************
\chapter{Related work}\label{ch:relatedWork}
%*****************************************
This chapter reviews research on \ac{lcw} and \ac{rag} advancements .
Both of these methods for knowledge processing and extraction are the subject of current \ac{ai} research and have experienced major advancements in recent time.
%
\section{LCW Models and Their Advancements}\label{sec:large_context}

Recent advancements in \ac{lm} architectures have led to a significant increase in context window sizes.
\ac{lm} have evolved from relatively short token limits (e.g., $2,048$ tokens in \textit{GPT-3}) to large windows exceeding $1{,}000{,}000$ tokens with the latest iterations of \ac{sota} \acp{llm} (e.g., \textit{Gemini $3$ Pro} \citet{Google2026Models}).
%
Large context windows are a desirable feature in \acp{llm}, but due to, for example, high fine-tuning costs, the scarcity of long text context windows is limited\citep{ding2024longrope}.
To avoid these resource-intensive approaches, current research is focusing on architectural modifications.

The overall goal is to extend the possible input sequence with minimal computational cost. 
Therefore, \citet{ding2024longrope} introduced \ac{longrope}, a \ac{rope} extension capable of scaling context windows beyond $2$ million tokens.
The performance of the models used is maintained in the process, while significantly improving long-text processing efficiency\citep{ding2024longrope}.
\citet{peng2024yarn} further emphasises that \ac{rope} effectively encodes positional information in transformer-based \ac{lm} to extend the initial smaller context window. 

Due to the remaining limitations of \ac{rope} in generalising beyond the initial context encodings, \citet{peng2024yarn} demonstrates a compute-efficient adaptation for long-context applications.
By applying \textit{YaRN} the tested \textit{LlaMa} models reached \ac{sota} performance  after fine-tuning on less than $\sim0.1\%$ of the original pre-training data.
Thus, significantly reducing the overhead  \citep{peng2024yarn}.
The fine-tuned models that were fine-tuned during the development and testing of \ac{yarn} have been made available online .
By making the \ac{yarn} technology open source, the context window extension has been reproduced with other language models of individual choice to a context length of $128,000$ token \citep{peng2024yarn}.
This is particularly relevant for \ac{rag} or \ac{qa} setups in resource-constrained environments.
%

Similarly, \citet{zhu2024poseefficientcontextwindow} proposed \ac{pose}.
A framework optimised for training long-context models with lower computational overhead. 
\ac{pose}-training decouples the training length from the target length.
Enabling an efficient context window extension during the \ac{pose}-training process that simulates long inputs using a fixed context window.
Experimental adaptation shows that \ac{pose} significantly reduces overhead compared to full-length fine-tuning \citep{zhu2024poseefficientcontextwindow}.
Therefore, possibly allowing implementation of less capable open-source \acp{llm} with thus extended context windows in projects with small hardware and time capabilities.
\subsection{Domain-Specific Possibilities and Evaluation}

In \citet{kinikogluI2025GPTGemini}, the performance of four of the most advanced models, \textit{ChatGPT-$4$o}, \textit{ChatGPT-o$1$}, \textit{Gemini $1.5$ Pro} and \textit{Gemini $2.0$ Advanced} was analysed on the Turkish Dental Specialist Examination (DUS) for 2020 and 2021.
All models demonstrated a high level of proficiency in answering the examination question with notable differences in their performance metrics.
%

The \textit{overall accuracy} of \textit{ChatGPT-o$1$} and \textit{Gemini $2.0$ Advanced} remained above $93$ for the exams evaluated that outperformed the less advanced models \textit{ChatGPT-4o} and \textit{Gemini 1.5 Pro}.
Further studies of the ability of \acp{llm} and their potential in medical education and assessment have shown similar results.
%
\citet{USMLE2023, chatgpt2023USMLE} demonstrated that \textit{ChatGPT} achieves pass scores on medical licensing examinations, such as the three US medical licensing examinations (USMLE), even without specific training. 
Therefore, \citet{kinikogluI2025GPTGemini} resumes that such \ac{sota} \acp{llm} can be adapted to create interactive and personalised learning experiences for students.
%

%
Despite these improvements, challenges remain in the use of \acp{llm} for example $-$ high computational costs for running the models, as well as potential inaccuracies, pre-trained biases and hallucinations during output generation.

Increased computational demands, potential retrieval inefficiencies, and memory constraints must be addressed to ensure effective application in educational and research environments. 
%

%
Considering the capabilities already available and ongoing developments, the integration of \acp{lm} with large context windows in domain-specific \ac{qa} offers significant opportunities - A fast and intuitive processing of large knowledge sources, whilst maintaining high accuracy in information retrieval.

The ability of such \acp{llm} to process entire large texts directly without the need for preprocessing or fine-tuning, represents a potential solution to the challenges outlined in \cref{sec:problem_statement}.
%

Research on \ac{lcw} focuses primarily on the following criteria:\\
Expanding the available context windows, and processing the input more efficiently and effectively.

The overall goal is to enable the most comprehensive context input possibly whilst minimising computational costs.

\section{Retrieval-Augmented Generation for QA}\label{sec:rag_for_qa}

\ac{rag} (\cref{sec:rag_prel}) as an alternative approach focuses per design on three things:\\
The dynamic and targeted \textbf{Retrieval} of relevant information, the subsequent \textbf{Augmentation} of knowledge of the respective \acp{lm} and finally the \textbf{Generation} of high quality output.
%
\subsection{Introduction of RAG}
Pre-trainig language models with billion-level parameters on huge amounts of training data as well as extending context windows have demonstrated impressive performance.
Despite vast capabilities, limitations remain: Factual inaccuracies \citep{augenstein2023factualitychallengeseralarge} and hallucinations \citep{fan2024surveyragmeetingllms}.
Their inability to expand and revise their memory \citep{lewis2020ragnlp}.
The demand for enormous computational resources to continuously develop and improve \acp{lm}.
%

%
In \citet{lewis2020ragnlp} a parametric-memory generation model in form of pre-trained \textit{seq2seq} transformer was extended with non-parametric memory, a dense vector index of Wikipedia.
The non-parametric memory is accessed via a pre-trained neural retriever.
\enquote{The retriever [\dots] provides latent documents conditioned on
the input, and the seq2seq model [\dots] then conditions on these latent documents together with the input to generate the output} \citep[p.2]{lewis2020ragnlp}
The resulting system is referred to as \textbf{retrieval-augmented generation (RAG)}.
%
\citet{lewis2020ragnlp}s approach differs from previous work on proposed architectures to enrich the system by the ability to access knowledge without training the \ac{lm} used for specific tasks from scratch.
Combining parametric and non-parametric memory with generation for knowledge-intensive tasks has achieved state-of-the-art results in the open domain \ac{qa}, as well as abstractive \ac{qa}.

It was also found that the \ac{rag} approach generates less hallucinations. Factually correct texts were generated more often compared to the non-augmented \ac{bart} model used \citep{lewis2020ragnlp}.
%
\subsection{Optimising RAG}
In the continuous progress and research of \ac{rag} since \citet{lewis2020ragnlp} several approaches to optimise the \ac{rag}-setup were researched.
\citet{izacard2022unsupervised, Karpukhin2020DensePR}, focused on improving the retriever that identifies and provides relevant information.
\citet{Agarwal2024ManyShotIL, liu2024lostmiddle} investigated the other side, the generator.
Research on both sides did approach the topic isolated with regard to their focus.
%

%
Existing work mainly proposed tuning the generator \ac{lm} used within \ac{rag}. \citet{Asai2023SelfRAGLT} explored and introduced a framework, \textit{SELF-RAG} to improve quality and factuality of the generating \ac{lm} through retrieval optimisation and \textit{self-reflection}.
During the development and application of \textit{SELF-RAG}, a generator \ac{lm} is trained on a corpus consisting of diverse text data.
That data is combined with reflection tokens and, depending on queries and context, additional relevant passages. 
The reflection tokens are inserted offline into the original corpus.
An additional critic model for an efficient reflection is therefore needed and applied, which reduces the overall overhead.
\textit{SELF-RAG} enables retrieval on demand.
\textit{SELF-RAG} processes passages in parallel and filters them through self-reflection.
Aspects of output quality, such as factuality, are also evaluated further.

Other works, such as \citet{lu2022QuarkCT, Korbak2023PretrainingLM} try to guide the \ac{lm} generation via use of general \textit{control tokens}.
Whereas \citet{Asai2023SelfRAGLT} enables prior evaluation of whether a retrieval is needed and then self-evaluate the quality of the generation.
The models developed by \citet{Asai2023SelfRAGLT} are able to outperform \textit{ChatGPT} (2023) in common Benchmarks like PopQA and ASQA (Rouge and MAUVE).

In applying the developed framework at the time available \ac{sota} \ac{lm} used within \textit{ChatGPT}, as well as the \textit{retrieval-augmented Llama2-chat}, were outperformed on open-domain \ac{qa}, as well as reasoning and fact verification tasks \citep{jin2024longcontextllmsmeetrag}.
%

%
The particular implementation of long-context- \ac{llm}s as generator in \ac{rag} and thus an optimal design of \ac{rag} remained under-explored \citep{jin2024longcontextllmsmeetrag}.
%
Therefore, \citet{jin2024longcontextllmsmeetrag} investigated the impact of the amount of retrieved passages on the generation step.
In particular, by leveraging the long-context windows of \ac{llm}.
\citet{jin2024longcontextllmsmeetrag} observed that an increasing number does not consistently improve performance.
At a certain point, the unconsidered increase of passages causes a decrease in retrieval quality caused by irrelevant information (\enquote{noise}).

\enquote{Simply providing more retrieved passages does not guarantee better outcome}\citep[p.2]{jin2024longcontextllmsmeetrag}.

Therefore, \citet{jin2024longcontextllmsmeetrag} proposes three methods: ($1$) \textbf{Retrieval reordering} based on retrieval scores.
This directly relates to the results of \citet{liu2024lostmiddle}.
Documents with higher scores placed at the beginning and end of the input sequence guide the respective \ac{lm} towards more relevant information \citep{jin2024longcontextllmsmeetrag}.\\
Further optimisation of the \acp{lm} can be achieved by ($2$) \textbf{Implicit robustness} fine-tuning and ($3$) \textbf{Implicit robustness fine-tuning}.
Hence, justifying the additional computational cost with the ability to \enquote{discern relevant information from noise within the retrieved context} \citep[p.2]{jin2024longcontextllmsmeetrag}.
%

%
Another work with particular relevance to resource-constrained solution approach provides \citet{Ma2023QueryRF}.
In the stage of the pre-retrieval process \textit{query optimisation} is used to make the original question more suitable for the retrieval task.
\citet{Ma2023QueryRF} extended the traditional \textit{retrieve-then-read} approach by adding a query rewriting step.
A small specified \ac{lm} acts as a rewriter.
Generating a rewritten query from the original input to improve retrieval techniques. 
\citet{Ma2023QueryRF} emphasises on the importance of query adaptation.
By the introduction of the \textit{Rewrite-Retrieve-Read} framework,  \citet{Ma2023QueryRF} offer a new perspective on enhancing retrieval-augmented \acp{llm}.
Tested on open-domain and multiple-choice \ac{qa} tasks, the framework application shows consistent performance improvements, demonstrating its effectiveness and scalability.
%
\section{Conclusion}
Significant advancements have been made in both \ac{rag} and \ac{lcllm} architectures that are experiencing continuous progress.
Due to this progress and a variety of requirements and specific and divers application scenarios, no universally optimal approach has emerged.
\citet{Li2024RetrievalAG} gives a detailed analysis of the advantages and disadvantage of \ac{rag} and \acp{llm} respectively.
The direct comparison shows that the use of the respective method in applications such as \ac{qa} is accompanied by compromises on both sides:
\begin{itemize}
    \item Intensive computing cost and time expenditure due to pre-training and fine-tuning (\ac{lcllm}).
    \item Hallucinations while generating output (\ac{lcllm}).
    \item Closed-source \ac{llm} models do not allow for further usage and extension (\ac{lcllm}).
    \item Potentially outdated knowledge due to pre-training on static data snapshots 
    \item No access to real-time updates when isolated outside an extending application (\ac{lcllm}).
    \item Lower accuracy for knowledge-intensive tasks (\ac{lcllm}).
    \item Lower overall response quality due to disjoint retrieval fragments (\ac{rag}).
    \item Retrieval errors (\ac{rag}).
    \item Dependency on external databases (\ac{rag}).
    \item Dependency on the quality and setup of the databases (\ac{rag}).
    \item Necessary adaptation of databases depending on the type of data
\end{itemize}
%
In response to these limitations, \citet{Li2024RetrievalAG} propose the hybrid approach via \textit{SELF-ROUTE}.
This framework attempts to dynamically select the most suitable method to optimise based on the specific requirements of the input query.
%
Overall \citet{Li2024RetrievalAG} confirm that, under optimal computational conditions, \acp{lcllm} generally outperform retrieval-based approaches.
However, \ac{rag} remains more efficient for short quires or limited computation scenarios.
%
\section{Perspective and Developmentens}

Both methods are constantly evolving.
Currently available applications combine and integrate various technologies and methods for information processing and generation.
Just four years have passed since the fascination of the \textit{ChatGPT} moment in $2022$ to the multi-modal, agentic appliactions of today.
These applications demonstrate that an isolated evaluation of models and methods is increasingly taking a back role in evaluations.

\textit{Google Gems}, for example, offers the creation of a custom \ac{ai} expert \citet{Google2025Gemsoverview}.
This expert can then assist to perform any requested task based on individual data and instructions and the underlying \ac{lm}.
\textit{Gems} combines three methods into one application - \textit{System instructions} (Defining the Persona), \textit{Large Context Window} (The Active Memory) and \textit{\ac{rag}} (targeted retrieval from provided datasets).
Similar applications exist.
They provide a basis for a future comparisons and evaluations of the available technologies and methods.

