%*****************************************
\chapter{Discussion}\label{ch:discussion}
%*****************************************

The results of this work show that leveraging the \ac{lcw} capabilties of \acp{llm} and the implementation of these models within \ac{rag} provide precise knowledge retrieval and high-quality output within \ac{qa}.
The method applied expectedly outperformed the continual pre-training method from \citet{Paul_Keller}.
The \acp{llm} tested here show better reasoning and \textit{Question Understanding} skills.
It has been demonstrated that the capability of large-context windows within \ac{qa} in domain-specific areas question the necessity of \ac{rag}.

However, using the models \enquote{out-of-the-box} and implementing similar settings as in \citet{Paul_Keller} introduces methodological trade-offs and limitations.
To potentially improve the performance of both method pipelines, \cref{ch:futurework} suggests several adaptations and extensions that can be investigated.

\section{Limitations of the Method Setup}

The method approaches for \ac{qa} using \acp{llm} with \citet{bb2} as knowledge base were particularly implemented to assess the models used \enquote{out-of-the-box-}.
The goal then was a direct comparison with \citet{Paul_Keller}, as well as between the models and methods.
Therefore, neither prompt engineering nor fine-tuning was deliberately applied.

This meets the premise of the experimental setup but, at the same time, clearly limits the potential performance of the models used.

Following the premise of creating a similar scenario as in \citet{bb2}, the token limit for the generated output was set to $512$.
The implementation showed that the models  generated answers during inference that extended the token limit set.
Although some models answered briefly without specific instructions, other models elaborated broadly, thereby generating a lot of \enquote{noise}.
Such behaviour was punished during evaluation, particularly by the \textit{Correctness} metric.

The \textit{OpenRouter} provider truncated the generated text on the server side.
As a result, the respective answers retrieved from the \ac{api} repsonse appeared to be incomplete.
In longer answers that contained larger amounts of noise, the truncation led to a significant loss of information.
  
Nevertheless, the \enquote{noise} provided information that could be evaluated as an understanding \textit{Understandability} of the question and the ability to provide an explanation \textit{Explainability}.
However, the content that remained after the truncation often did not contain a sufficient amount of valuable information in terms of facts.
This lead to a weak performance with regard to \textit{Correctness} was achieved.
%

In contrast, the \textit{Blablador} pipeline demonstrated the positive impact of implementing the token limit as an explicit instruction within the prompt.
This avoided hard cuts, preserved valuable information and enabled an undistorted evaluation of the generated output.
%

Furthermore, implementing the access to the \acp{lm} via \ac{api} providers solved the problem of local hardware limitations and the accessibility of \ac{llm}.
At the same time, it led to practical limitations and challenges.
The \ac{api} availability, latency, and request timeouts had a significant negative impact on the workflow using the models provided by \textit{Blablador}.

Data privacy remains an issue for the usage of \ac{lm} by commercial \ac{api} providers and \ac{ai} platforms in general.
The \enquote{black box} of server sided data processing can not provide any guarantees for the discreet and delicate handling of user data.
%
\section{Parametric Bias vs. Non-parametric Knowledge}

\acp{lm} are pre-trained on massive datasets, which pre-define their internal knowledge base.

Health Information Systems as a topic within medical informatics holds fundamental core definitions that are specific to and provided by \citet{bb2}.
These definitions may contradict the general literature or common usage on which the respective \ac{lm} has been pre-trained on.
The internal knowledge and the additional knowledge provided during inference are therefore in conflict.
Similarly to the findings of \citet{Paul_Keller}, this conflict is evident during the execution of the experiment.
%
\section{Compromises due to the methodology}

The implementation and subsequent comparison of the \ac{rag} and \ac{lcw} methods reveal their respective strengths and weaknesses.
Both in relation to \ac{qa} using a large and comprehensive knowledge base.
%
\subsection{LCW - Method}

The \ac{lcw} method is quite simple in its overall architectural setup.
The knowledge base is read and parsed to the respective \ac{lm} as a coherent text unit (string) within the prompt.

Due to the size of the parsed context, the models sometimes seem to be distracted and therefore unable to find and extract relevant knowledge with regard to the question.
If no relevant knowledge could be found in the knowledge base, the models would sometimes generated an answer based on its pre-trained, internal knowledge.
This potential misinformation is difficult to identify as such in the subsequent analysis.
The output usually provides plausible information.
However, in detail, it actually deviates from the ground truth provided by the knowledge base.
This type of answer is the most harmful, as students and practitioners are the ones who interpret it.
They lack the necessary corrective expert knowledge, which could result in them learning incorrect fundamentals. 
%
\subsection{RAG - Method}

Within the implemented \ac{rag} pipeline the retrieval accuracy was the most problematic parameter.
The first applied interquartile semantic chunking, could not return context that met the required similarity threshold.
This, again, resulted in the models answering based on their pre-trained knowledge or claiming that the provided context is insufficient to answer the question.
The latter is the outcome to be preferred in that scenario but was not constantly applied by the \acp{lm}.

The final hybrid chunking strategy partly solved the problem.

The thereby increased quantity of available chunks resulted in chunks meeting the threshold during retrieval.
However, these retrieved chunks occasionally provided distracting \enquote{noise}.
For example, a retrieved chunk contained an unrelated question involving \enquote{++} symbols.
The models misinterpreted the initial question to be related to the additional question.
The generated answer mainly referred to the second question and provided no useful content.

By definition, the \ac{rag} system consists of several components.
Each component plays an important role for the quality of output and can be individually enhanced and adapted.
Some of these available adaptions are described in \cref{ch:futurework}.
%
\section{Automated Evaluation}

The manual assessment and classification of the generated answers, as applied in \citet{Paul_Keller}, was replaced by an automated \enquote{LLM-as-a-judge} pipeline.
This implementation using the \textit{DeepEval} framework enabled an efficient and automatic evaluation process.

While this approach simplifies reproduction and guarantees scalability, it comes with some compromises.
%
\subsection{LLM-as-a-jduge}

The evaluating \ac{lm} instructed by a dedicated prompt often struggled to accurately separate the golden and generated answer into the required distinct semantic units.

The \ac{lm} as a judge and within \textit{Step $1$} also acting as the subanswer extractor, follows the defined instructions within the prompt while being influenced by its internal bias.

The underlying and instructing prompt cannot specify every possible nuance during the evaluation, which already limits the quality of result at this level.

Another factor of inconsistency during the evaluation was the formulation of the questions.
For example, the \texttt{single} question often contained formulations that indicated that at least $2$ subanswers were expected within the answer.
The fixed expected count of subanswers, however, required only $1$ subanswer.
F
\begin{lstlisting}[language=json, caption={Mismatch between question and expected subanswer count}, label={lst:disctracting_question}]
    {
    "transformed": "How would you describe the scope and tasks of the following activities of managing information systems: Daily monitoring of network availability and network failures",
    "num_answers": 1
    }
\end{lstlisting}
%
The evaluating \ac{lm} is instructed to directly integrate the initial question and the expected number of subanswers into the extraction process.
the question at \cref{lst:disctracting_question} indicates the answer to separated into at least $2$ contextual units.
One unit relates to the \enquote{scope}, while the other is concerned with the \enquote{tasks}.
The responsible judge now has to follow the instructed rule to separated distinct contextual units.
At the same time, it has to merge both units back into a single subanswer according to the expected count.
%
The issue becomes all the more siginificant the higher the expected number of subanswers.
%
All these possibilities and contextual nuances of the evaluation process must be defined in the prompt.
%
In conclusion, a subanswer extraction based on human assessment would, in the first place, simplify the process.
In the second place, it would improve the results of this step and the subsequent evaluation steps.

The \textit{Correctness} metric significantly depends on the quality of the output of \textit{Step $1$}.
Th necessary textit{Correctness} classification of the subanswers extracted is also significantly influenced by the instructing prompt.
The classification into \textit{correct}, \textit{incorrect}, \textit{not answered} is, in theory, quite distinct.
In practice, however, the interpretation of \textit{correct} and \textit{incorrect} is nuanced:
\begin{itemize}
    \item How should similar phrasing and incomplete matches in the subanswers by assessed?
    \item What if the subanswer contains knowledge that is factually correct but does not directly correspond to any golden true subanswer, yet is helpful in answering the question?
\end{itemize}
%
The guiding phrasing of the evaluation workflow posed a great challenge.
The evaluation should reflect the actual quality of the generated answer.
However, the instructing prompt had to find an appropriate level of complexity.

In conclusion, a classification by a human expert would result in a more reliable and probably more realistic representation of the model and method performance.

\section{Achievement of the set Goals}

The defined Goals (\cref{sec:goals}) could be achieved.
Goal G1 (\cref{def:goal 1}) was established by implementing a reproducible open-source application using Docker.
By using \ac{api} provider to use \ac{sota} \ac{llm} and running the application via Docker, the implemented setup is not restricted by resource-constraints.

Goal G2 (\cref{def:goal 2} and G3 (\cref{def:goal 3} were achieved, as the application automatically answered the domain-specific questions using \citet{bb2}
as knowledge base.
Finally, Goal G4 (\cref{def:goal 4} is met by implementing an automated evaluation framework that compared the models and methods according to the defined metrics and criteria.

The results demonstrate that the evaluated \acp{llm} are highly effective at accurately extracting, processing and reproducing the comprehensive domain-specific knowledge contained in \citet{bb2}.

The \ac{lcw} method achieved better results than the \ac{rag} method.
However, this does not allow for a general conclusion.
It must be interpreted in direct relation to the specific setup implemented here.



